Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-04 aaai-2018/khodadadi-mcdonald-aaai19.tex
"shrink figs, cutting down manuscript"

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_1
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

@@

\textcolor{red}{Our methodology is similar in flavor to~\cite{HallacPark2017}, with several fundamental differences. The main goal of Hallac et al. is to discover the time-varying structure of a network. To achieve this goal, they introduced two penalty terms into the objective function. One term penalizes the sum of the absolute values of the off-diagonal elements of the inverse covariance matrix. This term enforces sparsity on the estimated inverse covariance matrix and so helps in discovering which nodes in the network are connected. The other penalty term, penalizes differences in the estimated inverse covariance matrix from one time-step to the next. Our goal in the present work is to detect the temporal trend in the variance of each point in the network. The network in our problem is the grid of points over the earth and so the network's structure remains the same through time. The variance of each point, however changes. To capture these properties in our data using the objective function used in Hallac et al. (see Equation 2 in their paper), we need to have different temporal penalty terms for the diagonal and off-diagonal elements in the inverse covariance matrix. This results in unnecessary complication in the optimization problem. To overcome these difficulties, we introduced the new objective function \ref{eq:l1tf_var_st_mat}. This leads to a new optimization problem which we consider its solution next.}

\attn{Paragraph discussing relationship with estimating time-varying covariance}

@@

\begin{lemma}
  Let $f(x) = \sum_{k} x_{k} + y_{k}^2e^{-x_{k}}$ and $g(x) =
  \norm{x}_1$. Then, 
  \begin{align}
    [\prox_{\mu f}(u)]_k &= \mathscr{W}\bigg(\frac{y_k^2}{\mu}
    \exp\bigg(\frac{1-\mu u_k}{\mu}\bigg) \bigg) + \frac{1-\mu u_k}{\mu},\\
    \prox_{\rho g}(u) &= S_{\rho\lambda}(u)
  \end{align}
  where $\mathscr{W}(\cdot)$ is the \textit{Lambert W function}
  \cite{corless_lambertw_1996},  $[S_{\alpha}(u)]_k = \sign(u_k)(|u_k|
  -\alpha_k)_+$ and $(v)_+=v\vee 0$.
\end{lemma}
% \begin{proof}
%   If $f(x)=\sum_k f_k(x_k)$ then $[\prox_{\mu f}(x)]_k =
%   \prox_{\mu f_k}(u_k)$. So 
%   $[\prox_{\mu f}(u)]_k=\min_{x_k} \,\,
%   \mu\big(x_k+y_{k}^2e^{-x_{k}}\big)+\frac{1}{2}  (x_k-u_k)^2.$
%   Setting the derivative to 0 and solving for $u_k$ gives the
%   result. Similarly, $[\prox_{\rho g}(u)]_\ell=\rho
%   \lambda_\ell |z_\ell|+1/2(z_\ell-u_\ell)^2$. This is not differentiable,
%   but the solution must satisfy $\rho \cdot \lambda_\ell \cdot \partial
%   \big(|z_\ell| \big)=u_\ell-z_\ell$ where $\partial \big(|z_\ell| \big)$ is the
%   sub-differential of $|z_\ell|$. The solution is the soft-thresholding
%   operator $S_{\rho\lambda_\ell}(u_\ell)$.
% \end{proof}

@@

Before examining real data, we apply our model to some synthetic
data. This example was constructed to mimic the types of spatial and
temporal phenomena observable in typical climate data. We generate a
complete spatio temporal field wherein 
observations at all time steps and all locations are
independent Gaussian random variables with zero mean. However, the
variance of these random variables follows a smoothly varying function
in time and space given by the following parametric model:
\begin{align}
  \sigma^2(t,r,c) & =\sum_{k=1}^{K} W_k(t) \cdot \exp\bigg(
                    \frac{(r-r_k)^2+(c-c_k)^2} {2\sigma_k^2} \bigg)\\
  W_k(t) & =\alpha_k \cdot t + \exp(\sin(2\pi\omega_k t+\phi_k)) .
\label{eq:sourceVar}
\end{align}
In words, the variance at each time and location is computed as the
weighted sum of $K$ bell-shaped functions where the weights are
time-varying, consist of a linear trend $\alpha_k \cdot t$ and a
periodic term $\beta_k \cdot \sin(2\pi\omega_k t+\phi_k)$. The
bell-shaped functions impose spatial smoothness while the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in the real climate data. We simulated the
data on a 5 by 7 grid and for 780 time steps with $K=4$. This yields a
small enough problem so that it can be evaluated many times while
still mimicking important properties of climate data.
Specific parameter choices of the variance function are shown in
\autoref{tab:sim_params}.
For illustration, we also plot the
variance function for all locations at $t=25$ and $t=45$ in as well as
the variance across time at $(0,0)$ in \autoref{fig:true_var_spatial}.
\begin{table}[tb]
  \caption{Parameters used to simulate data.}
  \label{tab:sim_params}
  \begin{center}
    \begin{tabular}{ccccccc}
      \hline
      $s$ & $r_s$ & $c_s$ & $\sigma_s$ &$\alpha_s$ & $\omega_s$ & $\phi_s$\\
      \hline
      1 & 0 & 0 & 5 & 0.5 & 0.121 & 0 \\
      2 & 0 & 5 & 5 & 0.1 & 0.121 & 0 \\
      3 & 3 & 0 & 5 & -0.5 & 0.121 & $\pi/2$ \\
      4 & 3 & 5 & 5 & -0.1 & 0.121 & $\pi/2$ \\
      \hline
    \end{tabular}
  \end{center}
\end{table}

\begin{figure}[tb]
  \centering	
  \includegraphics[height=.15\textheight]{Figures/true_var_spatial}\\
  \includegraphics[height=.2\textheight]{Figures/true_fitted_var}
  \caption{Top: Variance function at $t=25$ (left) and $t=45$
    (right). Bottom: The true (orange) and estimated standard deviation 
    function at the location (0,0). The estimated values are
    obtained using linearized ADMM with $\lambda_s=0.1$ and two
    values of $\lambda_t$: $\lambda_t=5$ (blue) and
    $\lambda_t=100$ (green).} \label{fig:true_var_spatial}
\end{figure}

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. The
left panel of \autoref{fig:true_fitted_var} shows the true and the
estimated standard deviation at location (0,0) using $\lambda_s=0.1$
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green). As we can see,
larger than optimal value of $\lambda_t$ leads to estimated values
which are ``too smooth". The middle panel of \autoref{fig:true_fitted_var} shows the
convergence of both methods. Each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds. \attn{fix this discription}

\attn{Arash: Can you be more specific?}

\begin{figure}[tb]
  \centering	
  \includegraphics[height=.16\textheight]{Figures/convergence}  
  \includegraphics[height=.16\textheight]{Figures/modelComp_MAE}  
  \caption{Top: Convergence speed of linearized
    and consensus ADMM. Bottom: MAE for four models: our method: the
    proposed model with optimal values of $\lambda_t$ and $\lambda_s$,
    temporal: no spatial penalty, spatial: no temporal
    penalty.\attn{Fix these}} \label{fig:true_fitted_var}
\end{figure}

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which only imposes spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. \attn{What is this next sentence trying to say?}

\attn{that our model is better than GARCH in estimating the variance.}

Interestingly, the proposed model with optimal parameters 
outperforms GARCH(1,1) in estimating the true value of the variance.

@@

Consensus ADMM in~\autoref{sec:consOpt} is appropriate only if
we parallelize it over multiple machines, and it is significantly
slower on our simulated data, so we do not pursue it
further here. All the results reported in this section are obtained
using \autoref{alg:linADMM}. We
applied this algorithm to the northern hemisphere of the ERA-20C
dataset available from the European Center for Medium-Range
Weather Forecasts\footnote{\url{https://www.ecmwf.int}}. 
the 2 meter temperature measured daily at 12 p.m from January 1, 1960
to December 24, 2010.

@@

\attn{Can we redo these but with, say, Indianapolis instead of
  Bloomington?} Examination of the time-series alone demonstrates
strong differences 
between trend and cyclic behavior across spatial locations. One might
try to model the cycles by the summation of 
sinusoidal terms with different frequencies. However, for some
time-series this may need many terms to be included in the summation
to achieve a reasonable level of accuracy. In addition, such a model
cannot capture the non-stationarity in the cycles.

>>>>

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_t
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

@@

Our methodology is similar in flavor to~\cite{HallacPark2017} or
related work in~\citep{GibberdNelson2017,MontiHellyer2014}, but with
several fundamental differences. These papers aim to
discover the time-varying structure of a network. To achieve this
goal, they use Gaussian likelihood with unknown precision matrix and
introduce penalty terms which (1) encourage sparsity among the
off-diagonal elements and (2) discourage changes in the estimated inverse
covariance matrix from one time-step to the next.
Our goal in the present work is to detect the temporal trend in the variance of each
point in the network, but the network is known (corresponding to the grid
over the earth). The variance of each point, however changes. To modify the
objective function in~\cite[Eq.\ 2]{HallacPark2017}, we would
enforce complete sparsity on the off-diagonal elements (since
they are not estimated) and add a new penalty to enforce spatial
behavior across the diagonal elements. Thus, 
\eqref{eq:l1tf_var_st_mat} is not simply a special case of these
existing methods.

@@

\begin{lemma}
  Let $f(x) = \sum_{k} x_{k} + y_{k}^2e^{-x_{k}}$ and $g(x) =
  \norm{x}_1$. Then, 
  \begin{equation}
    \begin{aligned}
      [\prox_{\mu f}(u)]_k &= \mathscr{W}\bigg(\frac{y_k^2}{\mu}
      \exp\bigg(\frac{1-\mu u_k}{\mu}\bigg) \bigg) + \frac{1-\mu u_k}{\mu},\\
      \prox_{\rho g}(u) &= S_{\rho\lambda}(u)
    \end{aligned}
  \end{equation}
  where $\mathscr{W}(\cdot)$ is the \textit{Lambert W function}
  \cite{corless_lambertw_1996},  $[S_{\alpha}(u)]_k = \sign(u_k)(|u_k|
  -\alpha_k)_+$ and $(v)_+=v\vee 0$.
\end{lemma}
% \begin{proof}
%   If $f(x)=\sum_k f_k(x_k)$ then $[\prox_{\mu f}(x)]_k =
%   \prox_{\mu f_k}(u_k)$. So 
%   $[\prox_{\mu f}(u)]_k=\min_{x_k} \,\,
%   \mu\big(x_k+y_{k}^2e^{-x_{k}}\big)+\frac{1}{2}  (x_k-u_k)^2.$
%   Setting the derivative to 0 and solving for $u_k$ gives the
%   result. Similarly, $[\prox_{\rho g}(u)]_\ell=\rho
%   \lambda_\ell |z_\ell|+1/2(z_\ell-u_\ell)^2$. This is not differentiable,
%   but the solution must satisfy $\rho \cdot \lambda_\ell \cdot \partial
%   \big(|z_\ell| \big)=u_\ell-z_\ell$ where $\partial \big(|z_\ell| \big)$ is the
%   sub-differential of $|z_\ell|$. The solution is the soft-thresholding
%   operator $S_{\rho\lambda_\ell}(u_\ell)$.
% \end{proof}

@@

Before examining real data, we apply our model to some synthetic
data. This example was constructed to mimic the types of spatial and
temporal phenomena observable in typical climate data. We generate a
complete spatio temporal field wherein 
observations at all time steps and all locations are
independent Gaussian random variables with zero mean. However, the
variance of these random variables follows a smoothly varying function
in time and space given by the following parametric model:
\begin{align}
  \sigma^2(t,r,c) & =\sum_{k=1}^{K} W_k(t) \cdot \exp\bigg(
                    \frac{(r-r_k)^2+(c-c_k)^2} {2\sigma_k^2} \bigg)\\
  W_k(t) & =\alpha_k \cdot t + \exp(\sin(2\pi\omega_k t+\phi_k)) .
\label{eq:sourceVar}
\end{align}
The variance at each time and location is computed as the
weighted sum of $K$ bell-shaped functions where the weights are
time-varying, consist of a linear trend $\alpha_k \cdot t$ and a
periodic term $\beta_k \cdot \sin(2\pi\omega_k t+\phi_k)$. The
bell-shaped functions impose spatial smoothness while the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in the real climate data. We simulated the
data on a 5 by 7 grid and for 780 time steps with $K=4$. This yields a
small enough problem so that it can be evaluated many times while
still mimicking important properties of climate data.
Specific parameter choices of the variance function are shown in
\autoref{tab:sim_params}.
For illustration, we also plot the
variance function for all locations at $t=25$ and $t=45$
(\autoref{fig:true_var_spatial}, top panel) in as well as
the variance across time at $(0,0)$ (bottom panel, orange).
\begin{table}[tb]
  \caption{Parameters used to simulate data.}
  \label{tab:sim_params}
  \begin{center}
    \begin{tabular}{ccccccc}
      \hline
      $s$ & $r_s$ & $c_s$ & $\sigma_s$ &$\alpha_s$ & $\omega_s$ & $\phi_s$\\
      \hline
      1 & 0 & 0 & 5 & 0.5 & 0.121 & 0 \\
      2 & 0 & 5 & 5 & 0.1 & 0.121 & 0 \\
      3 & 3 & 0 & 5 & -0.5 & 0.121 & $\pi/2$ \\
      4 & 3 & 5 & 5 & -0.1 & 0.121 & $\pi/2$ \\
      \hline
    \end{tabular}
  \end{center}
\end{table} 
\begin{figure}[tb]
  \centering	
  \includegraphics[height=.1\textheight]{Figures/true_var_spatial}\\
  \includegraphics[height=.15\textheight]{Figures/true_fitted_var}
  \caption{Top: Variance function at $t=25$ (left) and $t=45$
    (right). Bottom: The true (orange) and estimated standard deviation 
    function at the location (0,0). The estimated values are
    obtained using linearized ADMM with $\lambda_s=0.1$ and two
    values of $\lambda_t$: $\lambda_t=5$ (blue) and
    $\lambda_t=100$ (green).} \label{fig:true_var_spatial}
\end{figure}

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. 
\begin{figure}[tb]
  \centering	
  \includegraphics[width=.46\columnwidth]{Figures/convergence}  
  \includegraphics[width=.46\columnwidth]{Figures/modelComp_MAE}  
  \caption{Left: Value of the objective function for linearized (orange)
    and consensus (blue) ADMM against iteration. Right: MAE for (1) our
    method with optimal values of $\lambda_t$ and $\lambda_s$ (2)
    temporal penalty only (3) spatial
    penalty only and (4) a GARCH(1,1).} 
  \label{fig:true_fitted_var}
\end{figure}
The
bottom panel of \autoref{fig:true_var_spatial} shows the true and the
estimated standard deviation at location (0,0) 
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green) ($\lambda_s=0.1$).
Larger values of $\lambda_t$ lead to estimated values
which are ``too smooth". The left panel of \autoref{fig:true_fitted_var} shows the
convergence of both methods as a function of iteration. It is
important to note that each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds. Thus, where the lines meet at
400 iterations requires about 4 seconds for the linearized method and
2 hours for the consensus method.

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which only imposes spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. We note that, using an algorithm akin
to~\citep{HallacPark2017} ignores the spatial component and

@@

Consensus ADMM in~\autoref{sec:consOpt} is appropriate when we can
easily parallelize it over multiple machines.  Otherwise, it is significantly
slower, so all the results reported in this section are obtained
using \autoref{alg:linADMM}. We
applied this algorithm to the Northern hemisphere of the ERA-20C
dataset available from the European Center for Medium-Range
Weather Forecasts\footnote{\url{https://www.ecmwf.int}}. We use 
the 2 meter temperature measured daily at 12 p.m from January 1, 1960
to December 24, 2010.

@@

Examination of the time-series alone demonstrates
strong differences 
between trend and cyclic behavior across spatial locations. One might
try to model the cycles by the summation of 
sinusoidal terms with different frequencies. However, for some
time-series this may need many terms to be included in the summation
to achieve a reasonable level of accuracy. In addition, such a model
cannot capture the non-stationarity in the cycles.
